TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-10-24 manuscript/manuscript.Rmd
Add comments from Silviu

Many studies have shown that genetically-driven complex phenotypes are often associated with functionally related genes that are more likely to form networks of interacting protein products [@Hartwell:1999aa; @Krauthammer:2004aa; @Huang:2006aa; @Lage:2007aa; @Cordell:2009aa; @Emily:2009aa; @Feldman:2008aa]. This observation has been verified systematically for a large number of diseases [@Menche:2015aa], thus confirming a fundamental hypothesis of the interactome-based approach to understanding human phenotype, namely that disease- and trait-associated genes tend to form interaction modules [@Goh:2007aa]. Such interaction modules may be viewed as connected subnetworks within the full interactome and may contain all molecular determinants of a certain phenotype. Consequently, methods to identify phenotype-associated networks of functionally-related genes have been developed [@Krauthammer:2004aa; @Cordell:2009aa; @Akula:2011aa].

Network properties, such as connectivity (aka degree distribution) [@Ghiassian:2015aa; @Lee:2008aa; @Barabasi:2004aa; @Feldman:2008aa], can be compared to better understand the relationships among phenotypes [@Vidal:2011aa; @Barabasi:2011aa; @Wang:2011aa]. Networks formed by phenotype-associated genes, or, more precisely, their protein-protein products, are thought to have high connectivity [@Huang:2006aa; @Krauthammer:2004aa; @Goh:2007aa]. This intuition reflects a well-known "guilt-by-association" principle that genes (or, more precisely, their products) that form an interaction network are more likely to share similar functions [@Farber:2013aa], or co-expression patterns [@Huang:2006aa; @Michalak:2008aa]. Although this view has been criticized [@Gillis:2012aa], several studies consistently observed phenotype-associated genes to be either connected as a single network or to participate in common phenotype-specific subnetworks [@Menche:2015aa; @Lee:2011aa; @Akula:2011aa; @Ideker:2008aa; @Iossifov:2008aa]. Therefore, high connectivity among phenotype-specific genes may indicate coherent molecular mechanisms that could be targeted therapeutically. In terms of graph theory, we are asking whether a phenotype-specific network is a community - a highly connected subgraph relatively well-separated from the rest of the network [@fortunato2010community]. However, the direct comparison of connectivity across phenotype-specific networks is hindered by the fact that different phenotypes are associated with different numbers of genes that form networks of different sizes and configurations.

@@

We created networks of phenotype-associated genes using PPI information from STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa] databases (see Methods). We opted for the use of three types of PPI data. First, STRING data filtered by interaction confidence score above 500 (middle of the bimodal distribution of the score, 0-1000 range, referred hereafter as “STRING filtered”) was used as the primary source of curated PPI data. This filtering step ensures that only high-confidence PPIs were selected. The advantage of using filtered data is the reliance on high-quality PPIs. The disadvantage is that the number of PPIs may be insufficient for forming phenotype-associated networks; consequently, phenotypes with genes without PPIs were removed from the analysis (see Methods) leavind 116 phenotypes that could be analyzed. Second, full STRING data was used to maximize the use of PPI information at the expense of potential noise (referred to hereafter as “STRING”). The advantage of using complete data is that more trait-associated genes will have PPI information and, hence, can be analyzed. The disadvantage, however, is that the results may be less reliable due to the presence of noisy PPIs having low confidence scores. Third, BioGrid PPI data, which has been curated to keep only high-confidence interactions, was used. The use of different data sources (STRING and BioGrid) and filtering (full or filtered STRING data) was intended to increase the generalizability of our conclusions.

@@

A typical network consists of nodes connected by edges [@Barabasi:2004aa]. In genomics, nodes typically represent genes and edges correspond to some measure of interactions, e.g., gene co-expression or interactions between protein products [@Goncalves:2012aa; @Wang:2012aa]. In terms of phenotypes, an intuitive expectation is that they will be represented by coherent networks of functionally related genes, similar gene expression profiles, shared genomic variants, higher PPI interactions, higher co-morbidity [@Zhang:2010ab; @Goh:2007aa; @Driel:2006aa; @Wang:2012aa; @Hamaneh:2015aa; @Menche:2015aa; @Jia:2011aa], reviewed in [@Dozmorov:2018aa; @Wang:2011aa]. In terms of graph theory, a coherent network is a community consisting of a group of nodes that are relatively highly connected to each other but sparsely connected to other nodes in the global network [@fortunato2010community; @Radicchi:2004aa].

@@

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed negative correlation of network size with the slopes of KEGG and random networks (Pearson Correlation Coefficient (PCC) -0.58/-0.86, respectively, Figure 2, Supplementary Table S2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

@@

We estimated coherence of 49 traits from five categories (Figure 3, Supplementary Table S3, Supplementary Figures S2-S3). Among anthropometric traits, Weight and Body Mass Index had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, the metabolic trait category had the highest overall coherence (mean coherence 0.86, Table 1), and these results were consistent when using other PPI databases (Supplementary Table S1). Several notable examples of low coherence include Obesity-related traits and Height (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the Intelligence category had low overall coherence (mean coherence = 0.41).

@@

## Coherence of most phenotypes significantly differs from random

Given the wide range of phenotype-specific coherence estimates, a natural question is whether they are significantly different from random coherence. We assessed the significance of coherence estimates using a permutation test (see Methods), individually for each PPI database. Permutation p-values obtained using Biogrid and STRING filtered databases were highly correlated (PCC = 0.64, Supplementary Figure S1), as would be expected for high-quality PPIs. Permutation p-values using STRING databases were predominantly significant (p-value < 0.001) due to a large number of PPIs that enabled phenotype-specific networks to be consistently different from random. Permutation p-values inversely correlated with coherence estimates (average PCC = -0.23), as would be expected for less coherent networks that resemble random networks. This dependency was especially pronounced for database-specific estimates, e.g., Biogrid coherence and p-value estimates correlated with PCC = -0.69 (STRING filtered PCC = -0.53, Supplementary Figure S1). These observations confirmed consistent coherence and p-value estimates when using different PPI databases.

The majority of phenotype-specific networks (~85%) were significantly coherent (STRING filtered p-value < 0.001, Supplementary Table S3). Among less significant phenotypes were "White blood cell count" (p-value = 0.085) and other blood count-related phenotypes, "Corneal astigmatism" (p-value = 0.04), "Major depressive disorder" (p-value = 0.03). When using Biogrid databases, "Educational attainment", "Myopia (pathological)", and "Celiac disease" were among the non-significant phenotypes (p-values 0.30, 0.20, 0.16, respectively). Interestingly, "Schizophrenia" network, despite being among the least coherent, was consistently significantly different from random (p-value < 0.0001). Overall, the significance estimation of coherence supports our observation that phenotype-specific networks in "Intelligence" and "Psychiatric disease" categories generally have low coherence that is frequently statistically insignificant. In summary, these results confirm our ability to identify coherence and of phenotype-specific networks with low, non-significant coherence.

@@

We developed a general-purpose method to measure and compare coherence estimates of the molecular mechanisms in networks of genes. Intuitively, gene networks with high internal and low external number of interactions (degrees) would be considered highly coherent; consequently, coherence is defined as the slope through the internal and external degree distribution scatterplot. We normalized coherences (slopes) to the range of high and random coherence, exemplified by MSigDb and random networks. To decouple coherence from its dependence on network size, we used MSigDb and random networks matched in size to phenotype-specific networks. Applied to the analysis of 116 disease- and trait-associated gene sets, we found that metabolic diseases and traits had overall high coherence, while psychiatric and cardiovascular diseases generally had low coherence. These results were consistent when using different PPI databases. Our method allows one to quantify and compare the estimated coherence of molecular mechanisms across phenotypes, and will only improve as more PPI information becomes available.

@@

The definition of coherence as the slope between internal vs. external degree distributions makes our measure dependent on the choice of PPI data [@Bajpai:2019aa]. Our limited knowledge of protein-protein interactions limits the coherence estimation to phenotypes that have a sufficient number of genes with PPI annotations, especially for phenotypes with low numbers of associated genes. Furthermore, networks formed by phenotype-associated genes should be sufficiently diverse to avoid cases in which all genes in a network have the same degree. As the different PPI databases vary significantly in the number of PPI annotations, some phenotype-associated gene sets have sufficient PPI annotations only when using one but not the other database. This problem is best illustrated by the BioGrid PPI database that has considerably fewer PPIs; hence, fewer phenotypes could be analyzed. <!--Differences in PPI databases may also lead to the presence of outliers that significantly alter coherence estimates. An example of such instability includes ESR2 and MOV10 genes in "Atrial Fibrillation" and "Blood Pressure", respectively, that have an unusually high number of external edges in Biogrid- but not STRING-based networks. Consequently, the corresponding networks appear incoherent/highly coherent, when using Biogrid and STRING PPI databases, respectively.--> Although our results show good correspondence between coherence estimates when using different PPI databases, care should be exercised when selecting the PPI database for network analyses [@Bajpai:2019aa].

@@

In addition to internal vs. external degree distributions, other gene-centric network metrics can be used to estimate coherence [@Leskovec:2010:ECA:1772690.1772755]. For example, the centrality measure has been used to demonstrate that highly central hub genes are functionally essential and highly conserved [@Khurana:2013aa]. A large study of PPIs (BioPlex) compared centrality distributions among several disease categories. Cancers and immunological disease networks tended to have high centrality, while nervous system, congenital, neonatal, and hereditary disease networks had low centrality [@Huttlin:2017aa]. Metabolic pathways were shown to have more duplicated copies of highly central genes making the networks tolerant to loss-of-function mutations [@Khurana:2013aa]. These observations of centrality measure parallel our estimation of network coherence. Our future work will incorporate centrality and other gene-centric network metrics into our framework to further refine our ability to compare the coherence of phenotype-associated molecular mechanisms.

>>>>

Many studies have shown that genetically-driven complex phenotypes are often associated with functionally related genes that are more likely to form networks of interacting protein products [@Hartwell:1999aa; @Krauthammer:2004aa; @Huang:2006aa; @Lage:2007aa; @Cordell:2009aa; @Emily:2009aa; @Feldman:2008aa]. This observation has been verified systematically for a large number of diseases [@Menche:2015aa], thus confirming a fundamental hypothesis of the interactome-based approach to understanding human phenotypes, namely that disease- and trait-associated genes tend to form interaction modules [@Goh:2007aa]. Such interaction modules may be viewed as connected subnetworks within the full interactome and may contain all molecular determinants of a certain phenotype. Consequently, methods to identify phenotype-associated networks of functionally-related genes have been developed [@Krauthammer:2004aa; @Cordell:2009aa; @Akula:2011aa].

Network properties, such as connectivity (aka degree distribution) [@Ghiassian:2015aa; @Lee:2008aa; @Barabasi:2004aa; @Feldman:2008aa], can be compared to better understand the relationships among phenotypes [@Vidal:2011aa; @Barabasi:2011aa; @Wang:2011aa]. Networks formed by interactions among phenotype-associated genes, or, more precisely, their protein products, are thought to have high connectivity [@Huang:2006aa; @Krauthammer:2004aa; @Goh:2007aa]. This intuition reflects a well-known "guilt-by-association" principle that genes (or, more precisely, their products) that form an interaction network are more likely to share similar functions [@Farber:2013aa], or co-expression patterns [@Huang:2006aa; @Michalak:2008aa]. Although this view has been criticized [@Gillis:2012aa], several studies consistently observed phenotype-associated genes to be either connected as a single network or to participate in common phenotype-specific subnetworks [@Menche:2015aa; @Lee:2011aa; @Akula:2011aa; @Ideker:2008aa; @Iossifov:2008aa]. Therefore, high connectivity among phenotype-specific genes may indicate coherent molecular mechanisms that could be targeted therapeutically. In terms of graph theory, we are asking whether a phenotype-specific network is a community - a highly connected subgraph relatively well-separated from the rest of the network [@fortunato2010community]. However, the direct comparison of connectivity across phenotype-specific networks is hindered by the fact that different phenotypes are associated with different numbers of genes forming networks of different sizes and configurations.

@@

We created networks of phenotype-associated genes using PPI information from STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa] databases (see Methods). We opted for the use of three types of PPI data. First, STRING data filtered by interaction confidence score above 500 (middle of the bimodal distribution of the score, 0-1000 range, referred hereafter as “STRING filtered”) was used as the primary source of curated PPI data. This filtering step ensures that only high-confidence PPIs were selected. The advantage of using filtered data is the reliance on high-quality PPIs. The disadvantage is that the number of PPIs may be insufficient for forming phenotype-associated networks; consequently, phenotypes with genes without PPIs were removed from the analysis (see Methods) leaving 116 phenotypes that could be analyzed. Second, full STRING data was used to maximize the use of PPI information at the expense of potential noise (referred to hereafter as “STRING”). The advantage of using complete data is that more trait-associated genes will have PPI information and, hence, can be analyzed. The disadvantage, however, is that the results may be less reliable due to the presence of noisy PPIs having low confidence scores. Third, BioGrid PPI data, which has been curated to keep only high-confidence interactions, was used. The use of different data sources (STRING and BioGrid) and filtering (full or filtered STRING data) was intended to increase the generalizability of our conclusions.

@@

A typical network consists of nodes connected by edges [@Barabasi:2004aa]. In genomics, nodes typically represent genes and edges correspond to some measure of interactions, e.g., gene co-expression or interactions between protein products [@Goncalves:2012aa; @Wang:2012aa]. In terms of phenotypes, an intuitive expectation is that they will be represented by coherent networks of functionally related genes, similar gene expression profiles, shared genomic variants, higher PPI interactions, and higher co-morbidity [@Zhang:2010ab; @Goh:2007aa; @Driel:2006aa; @Wang:2012aa; @Hamaneh:2015aa; @Menche:2015aa; @Jia:2011aa], reviewed in [@Dozmorov:2018aa; @Wang:2011aa]. In terms of graph theory, a coherent network is a community consisting of a group of nodes that are relatively highly connected to each other but sparsely connected to other nodes in the global network [@fortunato2010community; @Radicchi:2004aa].

@@

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed substantial negative correlation between network sizes and the slopes of KEGG and random networks (Pearson Correlation Coefficient (PCC) -0.58/-0.86, respectively, Figure 2, Supplementary Table S2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

@@

We estimated coherence of 49 traits from five categories (Figure 3, Supplementary Table S3, Supplementary Figures S2-S3). Among anthropometric traits, Weight and Body Mass Index had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, the metabolic trait category had the highest overall coherence (mean coherence 0.86, Table 1), and these results were consistent when using other PPI databases (Supplementary Table S1). Several notable examples of low coherence include Obesity-related traits, Height (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the Intelligence category had low overall coherence (mean coherence = 0.41).

@@

## Coherence of most phenotypes is significantly larger than random networks

Given the wide range of phenotype-specific coherence estimates, a natural question is whether they are significantly larger than the coherence of random networks. We assessed the significance of coherence estimates using a permutation test (see Methods), individually for each PPI database. Permutation p-values obtained using Biogrid and STRING filtered databases were highly correlated (PCC = 0.64, Supplementary Figure S1), as would be expected for high-quality PPIs. Permutation p-values using STRING databases were predominantly significant (p-value < 0.001) due to a large number of PPIs that enabled phenotype-specific networks to be consistently different from random. Permutation p-values inversely correlated with coherence estimates (average PCC = -0.23), as would be expected for less coherent networks that resemble random networks. This dependency was especially pronounced for database-specific estimates, e.g., Biogrid coherence and p-value estimates correlated with PCC = -0.69 (STRING filtered PCC = -0.53, Supplementary Figure S1). These observations confirmed that coherence and p-value estimates remained consistent when using different PPI databases.

The majority of phenotype-specific networks (~85%) were significantly more coherent than random networks (STRING filtered p-value < 0.001, Supplementary Table S3). Among less significant phenotypes were "White blood cell count" (p-value = 0.085) and other blood count-related phenotypes, "Corneal astigmatism" (p-value = 0.04), "Major depressive disorder" (p-value = 0.03). When using Biogrid databases, "Educational attainment", "Myopia (pathological)", and "Celiac disease" were among the non-significant phenotypes (p-values 0.30, 0.20, 0.16, respectively). Interestingly, "Schizophrenia" network, despite being among the least coherent, was consistently significantly different from random (p-value < 0.0001). Overall, the significance estimation of coherence supports our observation that phenotype-specific networks in "Intelligence" and "Psychiatric disease" categories generally have low coherence that is frequently statistically insignificant. In summary, these results confirm our ability to identify coherence of phenotype-specific networks and test their significance over random networks.

@@

We developed a general-purpose method to measure and compare the coherence estimates of the molecular mechanisms in networks of genes. Intuitively, gene networks with high internal and low external number of interactions (degrees) would be considered highly coherent; consequently, coherence is defined as the slope through the internal and external degree distribution scatterplot. We normalized coherences (slopes) to the range of high and random coherence, exemplified by MSigDb and random networks. To decouple coherence from its dependence on network size, we used MSigDb and random networks matched in size to phenotype-specific networks. Applied to the analysis of 116 disease- and trait-associated gene sets, we found that metabolic diseases and traits had overall high coherence, while psychiatric and cardiovascular diseases generally had lower coherence. These results were consistent when using different PPI databases. Our method allows one to quantify and compare the estimated coherence of molecular mechanisms across phenotypes, and will only improve as more PPI information becomes available.

@@

The definition of coherence as the slope between internal versus external degree distributions makes our measure dependent on the choice of PPI data [@Bajpai:2019aa]. Our limited knowledge of protein-protein interactions limits the coherence estimation to phenotypes that have a sufficient number of genes with PPI annotations, especially for phenotypes with low numbers of associated genes. Furthermore, networks formed by phenotype-associated genes should be sufficiently diverse to avoid cases in which all genes in a network have the same degree. As the different PPI databases vary significantly in the number of PPI annotations, some phenotype-associated gene sets have sufficient PPI annotations only when using one but not the other database. This problem is best illustrated by the BioGrid PPI database that has considerably fewer PPIs; hence, fewer phenotypes could be analyzed. <!--Differences in PPI databases may also lead to the presence of outliers that significantly alter coherence estimates. An example of such instability includes ESR2 and MOV10 genes in "Atrial Fibrillation" and "Blood Pressure", respectively, that have an unusually high number of external edges in Biogrid- but not STRING-based networks. Consequently, the corresponding networks appear incoherent/highly coherent, when using Biogrid and STRING PPI databases, respectively.--> Although our results show good correspondence between coherence estimates when using different PPI databases, care should be exercised when selecting the PPI database for network analyses [@Bajpai:2019aa].

@@

In addition to internal versus external degree distributions, other gene-centric network metrics can be used to estimate coherence [@Leskovec:2010:ECA:1772690.1772755]. For example, the centrality measure has been used to demonstrate that highly central hub genes are functionally essential and highly conserved [@Khurana:2013aa]. A large study of PPIs (BioPlex) compared centrality distributions among several disease categories. Cancers and immunological disease networks tended to have high centrality, while nervous system, congenital, neonatal, and hereditary disease networks had low centrality [@Huttlin:2017aa]. Metabolic pathways were shown to have more duplicated copies of highly central genes making the networks tolerant to loss-of-function mutations [@Khurana:2013aa]. These observations of centrality measure parallel our estimation of network coherence. Our future work will incorporate centrality and other gene-centric network metrics into our framework to further refine our ability to compare the coherence of phenotype-associated molecular mechanisms.
